Non-word repetition in Yélî Dnye: A small-scale study

### 2019-10-14 manuscript.Rmd
add quick comments on outline

- typical structure of NWR tasks: N of items
- complexity varies widely across tasks, sometimes within tasks on purpose
- some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
- calculation of frequency in Yêli is challenging bc low resource language
- so we only manipulated structural complexity, not probability
- previous work seems to avoid difficult sounds
- but we felt this was important to represent the language, so we also varied this factor

@@

1. Children are more accurate for mono-syllables than longer items
2. Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds
3. Children's accuracy increases with child age
4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the dominant language (we did not test any non-dominant language)
5. Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated
6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperfomr later-born children

@@

1. The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English. **Check for work on European languages that may have looked into this**
2. The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors. Therefore, this may constitute a pressure for children to have finer aduitory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory. **no work looking at consonants & vowels? no work looking at nasal vowels in particular?**
3. We had no reservations that children's accuracy increases with child age for the Yélî community as well.
4. Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from non-kin (or at least from people other than their mothers, who tend to be the non-native speakers). If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work. That said, one recent study shozs that most child-directed input in the first 2 years does come from the mother, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
5. In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment. This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly. However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
6. One main causal path between birth order and language development is via parental input (CITE). Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.

>>>>

- typical structure of NWR tasks: N of items
- complexity varies widely across tasks, sometimes within tasks on purpose
- some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
- calculation of frequency in Yêli is challenging bc low resource language **(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)**
- so we only manipulated structural complexity, not probability
- previous work seems to avoid difficult sounds
- but we felt this was important to represent the language, so we also varied this factor

@@

1. Children are more accurate for mono-syllables than longer items
2. Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds
3. Children's accuracy increases with child age
4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the dominant language (we did not test any non-dominant language) **(MC: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)**
5. Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated
6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperfomr later-born children

@@

1. The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English. **Check for work on European languages that may have looked into this**
2. The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors. Therefore, this may constitute a pressure for children to have finer aduitory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory. **no work looking at consonants & vowels? no work looking at nasal vowels in particular?**
3. We had no reservations that children's accuracy increases with child age for the Yélî community as well.
4. Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from non-kin **(MC: definitely almost all kin; 5k of them)** (or at least from people other than their mothers, who tend to be the non-native speakers). If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work **(MC: maybe it'd be more useful to try and estimate the number of non-Y moms who live in each hamlet where testing is taking place?)**. That said, one recent study shozs that most child-directed input in the first 2 years does come from the mother **(MC: i can get you actual estimates for this for YD; let's go with that instead!)**, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
5. In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment. This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly. However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
6. One main causal path between birth order and language development is via parental input (CITE). Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.
